Power and Pitfalls of Generic Smart Contracts

### 2022-03-01 cluster_computing/001_abstract.tex
Reorganized figures

Generics are a powerful feature of programming languages that allows one
to write highly reusable code.
%
More specifically, they are based on the use of type placeholders in order
to produce parametrized code, that can be instantiated for each
concrete type provided for them. 
%
In many programming languages, such as Java, they are implemented by
\emph{erasure}\ie replaced by their upper bound type during compilation into bytecode.
This is safe at source level, since the compiler takes care of checking that
types are correct, before erasure. However, the types of the generated bytecode
are erased and consequently weaker. In a permissionless blockchain, where
every user can call the bytecode of smart contracts installed by other users,
these weaker types pose a risk of attack.
%
Namely, this paper originated from a real security issue that we found
while using generics for writing
smart contracts that implement
\emph{shared entities} (such as a company shared by its shareholders),
for the Hotmoka blockchain, in a subset of Java called Takamaka.
%
The considered case study is particularly important since
the validators' set of the blockchain itself is
a special case of shared entities.
The analysis shows that the power of generics comes at the risk of
allowing a validator of an illegal type, that exploits the work of another
validator in order to earn validation rewards, an actual attack
to the validators' set. This paper proposes a solution
that forces the compiler to generate more precise type information than
that generated by erasure.

>>>>

Generics are a powerful feature of programming languages that allows one
to write highly reusable code.
%
More specifically, they are based on the use of type placeholders in order
to produce parametrized code, that can be instantiated for each
concrete type provided for the placeholders.
%
In many programming languages, such as Java, generics are implemented by
\emph{erasure}\ie replaced by their upper bound type during compilation into bytecode.
This is safe at source level, since the compiler takes care of checking that
types are correct, before erasure. However, the types of the generated bytecode
are erased and consequently weaker. In a permissionless blockchain, where
every user can call the bytecode of smart contracts installed by other users,
these weaker types pose a risk of attack.
%
Namely, this paper originated from a real security issue that we found
while using generics for writing
smart contracts that implement
\emph{shared entities} (such as a company shared by its shareholders),
for the Hotmoka blockchain, in a subset of Java called Takamaka.
%
The considered case study is particularly important since
the validators' set of the blockchain itself is
a special case of shared entity.
The analysis shows that the power of generics comes at the risk of
allowing the intrusion of
a validator of an illegal type, that exploits the work of another
validator in order to earn validation rewards, an actual attack
to the validators' set. This paper proposes a solution
that forces the compiler to generate more precise type information than
that generated by erasure.

### 2022-03-01 cluster_computing/010_introduction.tex
Reorganized figures

Strong typing is one of the reasons that push towards the adoption
of \emph{traditional} programming languages for smart contracts. For instance,
the Cosmos blockchain~\cite{cosmos} uses Go. The
Hotmoka blockchain~\cite{hotmoka} uses a subset of Java
for smart contracts, called Takamaka~\cite{Spoto19,Spoto20}.
Hyperledger~\cite{hyperldeger} allows Go and Java.
Another reason is the availability of modern
language features, that are missing in Solidity,
such as \emph{generics}\ie the possibility of using
type variables. Generics are a powerful and very useful facility for programming
smart contracts, since they allow one to personalize the behaviour of such contracts and partially overcome their inherent incompleteness~\cite{ebp}. In Java source code, generics are strongly typed, if no \emph{unchecked operations}
are used~\cite{NaftalinW06}, as it will always be the case in this paper.
However, generics might have security issue at the level of compiled Java code and this paper originated from a
real issue that we found in our code.

The contribution of this paper is to show a real-life
use of generics for an actual smart contract contained in the support
library of the Takamaka language, and to demonstrate that a na\"{i}ve use
of Java generics can lead to a code security vulnerability which
allows an attacker to earn money by exploiting someone else's work, with both economical and legal side effects.
This paper will provide a fix to that specific issue,
by proposing a re-engineering of the code that forces the compiler to generate defensive checks.
More generally, this paper can be useful for the definition of
bytecode languages for future smart contract languages, by
learning from the weaknesses of Java bytecode.

The remainder of this paper is organized as follows.
Sec.~\ref{sec:java_generics} discusses the management of generics in Java.
Sec.~\ref{sec:takamaka} presents the basic notions about the Takamaka language for smart contracts in Java.
Sec.~\ref{sec:shared_entities} shows our real-life Java smart
contracts shared entities, that use generic types.
Sec.~\ref{sec:validators} shows the instantiation of the shared entities to implement the validators' set
of a proof of stake blockchain.
Sec.~\ref{sec:attack} shows that a na\"{i}ve
deployment of a subclass of the validators' set leads to a code vulnerability.
Sec.~\ref{sec:fix} presents a fix to that vulnerability.
Sec.~\ref{sec:related_work} discusses some related work.
Sec.~\ref{sec:conclusion} concludes.

>>>>

Strong typing is one of the reasons that push towards the adoption
of \emph{traditional} programming languages for smart contracts. For instance,
the Cosmos blockchain~\cite{cosmos} uses Go. The
Hotmoka blockchain~\cite{hotmoka} uses a subset of Java
for smart contracts, called Takamaka~\cite{Spoto19,Spoto20}.
Hyperledger~\cite{hyperldeger} allows Go and Java.
Another reason is the availability of modern
language features, that are missing in Solidity,
such as \emph{generics}\ie the possibility of using
type variables. Generics are a powerful and very useful facility for programming
smart contracts, since they allow one to personalize the behaviour of such contracts and partially overcome their inherent incompleteness~\cite{ebp}. In Java source code, generics are strongly typed, if no \emph{unchecked operations}
are used~\cite{NaftalinW06}, as it will always be the case in this paper.
However, generics might have security issues
at the level of compiled Java code and this paper originated from a
real issue that we found in our code.

The contribution of this paper is to show a real-life
use of generics for an actual smart contract contained in the support
library of the Takamaka language, and to demonstrate that a na\"{i}ve use
of Java generics can lead to a code security vulnerability that
allows an attacker to earn money by exploiting someone else's work, with both economical and legal side effects.
This paper will provide a fix to that specific issue,
by proposing a re-engineering of the code that forces the compiler to generate defensive checks.
More generally, this paper can be useful for the definition of
bytecode languages for future smart contract languages, by
learning from the weaknesses of Java bytecode.

The remainder of this paper is organized as follows.
Sec.~\ref{sec:java_generics} discusses the management of generics in Java.
Sec.~\ref{sec:takamaka} presents the basic notions about the Takamaka language for smart contracts in Java.
Sec.~\ref{sec:shared_entities} shows our real-life Java smart
contracts for shared entities, that use generic types.
Sec.~\ref{sec:validators} shows the instantiation of the shared entities to implement the validators' set
of a proof of stake blockchain.
Sec.~\ref{sec:attack} shows that a na\"{i}ve
deployment of a subclass of the validators' set leads to a code vulnerability.
Sec.~\ref{sec:fix} presents a fix to that vulnerability.
Sec.~\ref{sec:related_work} discusses some related work.
Sec.~\ref{sec:conclusion} concludes.

This paper is a revised and extended version of~\cite{BeniniGMS21}.
In comparison to that paper, Sec.~\ref{sec:takamaka} and
Sec.~\ref{sec:validators} are new; eight figures have been added;
all other sections have been expanded.

### 2022-03-01 cluster_computing/015_generics_in_java.tex
Reorganized figures

There exists two common ways to implement generics in a programming language,
that are often described in literature as \emph{heterogeneous}
and \emph{homogeneous}~\cite{generics_categories}.
In the heterogeneous approach, the code is duplicated and specialized for each instance
of the generic parameters; this is the approach adopted by C++ \emph{templates}.
Conversely, the homogeneous approach is that provided by Java and .Net; in this case,
only one instance of the code is maintained and shared by all generic instances.
This implementation is based on the type \emph{erasure} mechanism, where the generic parameter
is replaced by the upwards bound of each instance, mostly often \<Object>.
Even though the heterogeneous approach is the safest, it is rarely applied in particular
in resource-constrained applications, because the code size may dramatically increase
because of duplication~\cite{generics_embedded_systems}. For code in blockchain,
the heterogeneous approach obliges one to reinstall all instantiations of the generic code,
with extra costs of gas, which makes it impractical.
Conversely, the homogeneous approach ensures a smaller consumption of resources.

@@

Erasure weakens the type information of the compiled code. It is the responsibility of the
compiler to guarantee that types are still respected, in all implementations of \<SharedEntity>.
In Java, the compiler guarantees type correctness and the Java language remains strongly-typed,
also in the presence of generic types,
if no \emph{unchecked operations} are performed~\cite{NaftalinW06} (such as casts to generic types,
that are unchecked for a limitation of the Java bytecode).
However, this guarantee applies to Java source code compiled by the Java compiler,
not to bytecode that can be generated manually, in order to attack class \<SharedEntity>,
as we will show later.

>>>>

There exist two common ways to implement generics in a programming language,
that are often described in literature as \emph{heterogeneous}
and \emph{homogeneous}~\cite{generics_categories}.
In the heterogeneous approach, the code is duplicated and specialized for each instance
of the generic parameters; this is the approach adopted by C++ \emph{templates}.
Conversely, the homogeneous approach is that provided by Java and .Net; in this case,
only one instance of the code is maintained and shared by all generic instances.
This implementation is based on the type \emph{erasure} mechanism, where the generic parameter
is replaced by the upwards bound of each instance, mostly often \<Object>.
Even though the heterogeneous approach is the safest, it is rarely applied, in particular
in resource-constrained applications, because the code size may dramatically increase
as a consequence of duplication~\cite{generics_embedded_systems}. For code in blockchain,
the heterogeneous approach obliges one to reinstall all instantiations of the generic code,
with extra costs of gas, which makes it impractical.
Conversely, the homogeneous approach ensures a smaller consumption of resources.

@@

Erasure weakens the type information of the compiled code. It is the responsibility of the
compiler to guarantee that types are still respected, in all implementations of \<SharedEntity>.
In Java, the compiler guarantees type correctness and the Java language remains strongly-typed,
also in the presence of generic types,
if no \emph{unchecked operations} are performed~\cite{NaftalinW06} (such as casts to generic types,
that are unchecked for a limitation of the Java bytecode).
However, this guarantee applies to Java source code compiled by the Java compiler,
not to bytecode that can be generated manually, in order to attack instance of
the \<SharedEntity> class, as we will show later.

### 2022-03-01 cluster_computing/017_takamaka.tex
figures
Reorganized figures

This section gives a short introduction to the Takamaka subset of
Java that we use for writing smart contracts in this paper.
This language has been introduced in~\cite{Spoto19}. A full tutorial
is available online, as part of the documentation of the Hotmoka
blockchain that runs smart contracts written
in Takamaka~\cite{hotmoka_repository}. This section introduces only
the essential notions that are needed to understand the subsequent sections.
The hierarchy of the classes described in this section is in Fig.~\ref{fig:hierarchy-entities}.
In the following we will report a simplified presentation of the code of some
of such classes. The full code is in the Github repository of Hotmoka \footnote{\url{https://github.com/Hotmoka}}.

@@

Takamaka implements objects persisted in blockchain as subclasses of the
\<io.takamaka.code.lang.Storage> class.
This is the main difference with other attemps at using Java for writing smart contracts:
the programmer does not code the serialization and deserialization of objects into
a \emph{keeper} or a key/value map, but simply extends \<Storage> and objects get
persisted automatically \emph{out of magic}. In this sense, Takamaka follows the apporach of
Solidity, but using Java.

@@

  @Override public @View final String id() {
    return id;
  }
}
    \end{lstlisting}
  \end{center}
  \caption{The account of a validator and its specialization for a Hotmoka blockchain based on Tendermint.}\label{fig:validator}
\end{figure}

The \<io.takamaka.code.lang.Contract> class implements objects that can be
persisted in blockchain \emph{and} have a balance. Therefore, they can receive and provide payments.
Their balance is available through a \<balance> method. This is a \<@View> method, meaning
that it can be called without paying gas (the measure of execution cost), since such methods
cannot have side-effects and consequently do not modify the storage of the blockchain.
Payments can be received only through methods annotated as \<@Payable>.
The \<Contract> superclass has no such methods, but subclasses may have.
For instance, its \<io.takamaka.code.lang.PayableContract> subclass has
a method \<receive> to receive payments from its caller. Many methods (inclusing all
\<@Payable> methods) need to identify their caller. This is done by adding the
\<@FromContract> annotation, that guarantees that the caller is a contract, that is available
inside the method as \<caller()>.

\begin{figure}[ht]
\centering
\includegraphics[width=.9\linewidth]{figures/takamaka_payable_contract}
\caption{}
\label{figure.takamaka_payable_contract}
\end{figure}

Neither the Takamaka language nor the Hotmoka blockchain dictate a specific consensus mechanism.
Both proof of work and proof of stake can be used, for instance. In particular, if proof of stake
is used, then each validator node of the blockchain must specify a
\<io.takamaka.code.governance.Validator> object, that plays the role of the
banking account where the validation rewards of the node get accumulated (see Fig.~\ref{fig:validator}).
It is a special externally owned account, with an extra \<id> method that provides the
identifier of the validator node inside the blockchain network. This identifier depends from the
specific network.
For instance, the subclass
\<io.takamaka.code.governance.tendermint.TendermintED25519Validator>
implements \<id> as for the Tendermint blockchain engine, that is,
as the first $40$ hexadecimal digits of the sha256 digest of the Base64-encoded public key.

>>>>

This section gives a short introduction to the Takamaka subset of
Java that we use for writing smart contracts in this paper.
This language has been introduced in~\cite{Spoto19}. A full tutorial
is available online, as part of the documentation of the Hotmoka
blockchain that runs smart contracts written
in Takamaka~\cite{hotmoka_repository}. This section introduces only
the essential notions that are needed to understand the subsequent sections.
The hierarchy of the classes described in this section is in Fig.~\ref{fig:hierarchy-entities}.
In the following we will report a simplified presentation of the code of some
of such classes. The full code is in the Github repository of Hotmoka~\cite{hotmoka_repository}.

@@

Takamaka implements objects persisted in blockchain as subclasses of the
\<io.takamaka.code.lang.Storage> class.
This is the main difference with other attempts at using Java for writing smart contracts:
the programmer does not code the serialization and deserialization of objects into
a \emph{keeper} or a key/value map, but simply extends \<Storage> and objects get
persisted automatically \emph{out of magic}. In this sense, Takamaka follows the approach of
Solidity, but using Java.

The \<io.takamaka.code.lang.Contract> class implements objects that can be
persisted in blockchain \emph{and} have a balance. Therefore, they can receive and provide payments.
Their balance is available through a \<balance> method. This is a \<@View> method, meaning
that it can be called without paying gas (the measure of execution cost), since such methods
cannot have side-effects and consequently do not modify the storage of the blockchain.
Payments can be received only through methods annotated as \<@Payable>.
The \<Contract> superclass has no such methods, but subclasses may have.
For instance, its \<io.takamaka.code.lang.PayableContract> subclass has
a method \<receive> to receive payments from its caller. Many methods (inclusing all
\<@Payable> methods) need to identify their caller. This is done by adding the
\<@FromContract> annotation, that guarantees that the caller is a contract, available
inside the method as \<caller()>.

\begin{figure}[ht]
\centering
\includegraphics[width=.9\linewidth]{figures/takamaka_payable_contract}
\caption{Exemplification of the Takamaka persistent objects stored in the blockchain.}
\label{figure.takamaka_payable_contract}
\end{figure}

@@

  @Override @View public final String id() {
    return id;
  }
}
    \end{lstlisting}
  \end{center}
  \caption{The account of a validator and its specialization for a Hotmoka blockchain based on Tendermint.}\label{fig:validator}
\end{figure}

Neither the Takamaka language nor the Hotmoka blockchain dictate a specific consensus mechanism.
Both proof of work and proof of stake can be used, for instance. In particular, if proof of stake
is used, then each validator node of the blockchain must specify a
\<io.takamaka.code.governance.Validator> object, that plays the role of the
banking account where the validation rewards of the node get accumulated (see Fig.~\ref{fig:validator}).
It is a special externally owned account, with an extra \<id> method that provides the
identifier of the validator node inside the blockchain network. This identifier depends from the
specific network.
For instance, the subclass
\<io.takamaka.code.governance.tendermint.TendermintED25519Validator>
implements \<id> as for the Tendermint blockchain engine, that is,
as the first $40$ hexadecimal digits of the sha256 digest of the Base64-encoded public key
(see its code in Fig.~\ref{fig:validator}).

### 2022-03-01 cluster_computing/020_shared_entities_with_generics.tex
Reorganized figures

@@

  public SimpleSharedEntity
      (S shareholder, BigInteger share) {
    addShares(shareholder, share);
  }

@@

The \<SharedEntity> subinterface adds methods for transfer of shares
(see Fig.~\ref{fig:shared_entity}).
It includes an inner class \<Offer> that models sale offers:
it specifies who is the seller of the shares,
how many shares are being sold, the requested price and the expiration of the offer.
Method \<isOngoing> checks if an offer has not expired yet.
Implementations can subclass \<Offer> if they need more specific offers.
Offers can be placed on sale
by calling the \<place> method with a sale \<offer>.
This method is annotated as \<@FromContract> since the caller must be
identified (or otherwise anybody could sell the shares of anybody else) and
as \<@Payable> so that implementations can require
to pay a \<ticket> to place shares on sale.
The sale offer is passed as a parameter to \<place>, hence it must have been created before calling that method.
The set of all sale offers is available through \<getOffers>. Method \<sharesOnSale> yields the
cumulative number of shares on sale for a given shareholder.
Who wants to buy shares calls method \<accept> with the accepted \<offer>
and with itself as \<buyer> (the reason will be explained soon)
and becomes a new shareholder or increases
its cumulative number of shares (if it was a shareholder already).
Also this method is \<@Payable>, since its caller must pay \<ticket> $\ge$ \<offer.cost>
coins to the seller.
This means that shareholders must be able to receive payments and that
is why \<S extends PayableContract>: only \<PayableContract>s are guaranteed to have a
\<receive> method in Takamaka.

>>>>

The \<SharedEntity> subinterface adds methods for transfer of shares
(see Fig.~\ref{fig:shared_entity}).
It includes an inner class \<Offer> that models sale offers:
it specifies who is the seller of the shares,
how many shares are being sold, the requested price and the expiration of the offer.
Method \<isOngoing> checks if an offer has not expired yet.
Implementations can subclass \<Offer> if they need more specific offers.
Offers can be placed on sale
by calling the \<place> method with a sale \<offer>
(see Fig.~\ref{figure.shared_entity}).
This method is annotated as \<@FromContract> since the caller must be
identified (or otherwise anybody could sell the shares of anybody else) and
as \<@Payable> so that implementations can require
to pay a \<ticket> to place shares on sale.
The sale offer is passed as a parameter to \<place>, hence it must have been created before calling that method.
The set of all sale offers is available through \<getOffers>. Method \<sharesOnSale> yields the
cumulative number of shares on sale for a given shareholder.
Who wants to buy shares calls method \<accept> with the accepted \<offer>
and with itself as \<buyer> (the reason will be explained soon)
and becomes a new shareholder or increases
its cumulative number of shares (if it was a shareholder already).
Also this method is \<@Payable>, since its caller must pay \<ticket> $\ge$ \<offer.cost>
coins to the seller.
This means that shareholders must be able to receive payments and that
is why \<S extends PayableContract>: only \<PayableContract>s are guaranteed to have a
\<receive> method in Takamaka.

@@

  public SimpleSharedEntity(S shareholder, BigInteger share) {
    addShares(shareholder, share);
  }

@@

### 2022-03-01 cluster_computing/025_validators.tex
figures
Reorganized figures

The Hotmoka blockchain is built over Tendermint~\cite{Kwon14}, a
generic engine for replicating an application over a network of nodes. In our case,
the application is the executor of smart contracts in Java, such as that in
Fig.~\ref{fig:simple_shared_entity}. Tendermint is based on a proof of stake
consensus, which means that a selected dynamic subset of the nodes is in charge of
validating the transactions and voting their acceptance. As we have said,
Hotmoka models validator nodes as \<Validator> objects, that are externally owned accounts
with an extra identifier. In the specific case of a Hotmoka blockchain built over Tendermint,
validators are \<TendermintED25519Validator> objects whose
identifier is derived from their ed25519 public key (see Fig.~\ref{fig:validator}).
This identifier is public information, reported in the blocks or easily eavesdropped.
Tendermint applications can implement their own
policy for rewarding or changing the validators' set dynamically.

The \<AbstractValidators> class implements the validators' set
and the distribution of the reward and is a subclass of \<SimpleSharedEntity>
(see Fig.~\ref{fig:validators}).
Shares are voting power in this case.
Its subclass \<TendermintValidators> restricts the type of the validators
to be \<TendermintED25519Validator>.
At each block committed, Hotmoka calls the \<reward> method of \<Validators>
in order to reward the validators that behaved correctly
and slash those that misbehaved, possibly removing them from the validators' set.
They are specified by two strings
that contain the identifiers of the validators, as provided by the underlying
Tendermint engine.

@@

public abstract class AbstractValidators
    <V extends Validator>
    extends SimpleSharedEntity<V,Offer<V>> {

  public AbstractValidators
      (V validator, BigInteger power) {
    super(validator, power);
  }

@@

>>>>

The Hotmoka blockchain is built over Tendermint~\cite{Kwon14}, a
generic engine for replicating an application over a network of nodes. In our case,
the application is an executor of smart contracts in Java, such as that in
Fig.~\ref{fig:simple_shared_entity}. Tendermint is based on a proof of stake
consensus, which means that a selected dynamic subset of the nodes is in charge of
validating the transactions and voting their acceptance. 
As we have said,
Hotmoka models validator nodes as \<Validator> objects, that are externally owned accounts
with an extra identifier. In the specific case of a Hotmoka blockchain built over Tendermint,
validators are \<TendermintED25519Validator> objects whose
identifier is derived from their ed25519 public key (see Fig.~\ref{fig:validator}).
This identifier is public information, reported in the blocks or easily eavesdropped.
Tendermint applications can implement their own
policy for rewarding or changing the validators' set dynamically.

@@

public abstract class AbstractValidators<V extends Validator>
    extends SimpleSharedEntity<V,Offer<V>> {

  public AbstractValidators(V validator, BigInteger power) {
    super(validator, power);
  }

@@

\begin{figure}[ht]
\centering
\includegraphics[width=.9\linewidth]{figures/takamaka_validators}
\caption{Hierarchy of classes for implementing Hotmoka Validators.}
\label{figure.takamaka_validators}
\end{figure}

The \<AbstractValidators> class implements the validators' set
and the distribution of the reward and is a subclass of \<SimpleSharedEntity>
(see Fig.~\ref{fig:validators} and Fig.~\ref{figure.takamaka_validators}).
Shares are voting power in this case.
Its subclass \<TendermintValidators> restricts the type of the validators
to be \<TendermintED25519Validator>.
At each block committed, Hotmoka calls the \<reward> method of \<Validators>
in order to reward the validators that behaved correctly
and slash those that misbehaved, possibly removing them from the validators' set.
They are specified by two strings
that contain the identifiers of the validators, as provided by the underlying
Tendermint engine.

### 2022-03-01 cluster_computing/030_generics_attack.tex
figures
Reorganized figures

In general, the problem arises whenever a method declares a formal parameter of generic type,
such as \<buyer> in method \<accept> of Fig.~\ref{fig:shared_entity}. Next section shows
the actual significance of the consequent security risk.

In particular, it is important that the \emph{Consistency of Shareholders} property holds
for the subclass \<TendermintValidators>: its shareholders
must be \<TendermintED25519Validator>s (as declared in the generic
signature of \<TendermintValidators> in Fig.~\ref{fig:validators}) that enforce a
match between their public key, that identifies who can spend the rewards sent to the validator,
and their Tendermint identifier, that identifies which node of the blockchain
must do the validation work (see how the constructor initializes \<this.id> in
Fig.~\ref{fig:validator}).
If it were possible to add a shareholder of another
type \<Attacker>, the code of \<Attacker> could decouple the node identifier from its
public key (see Fig.~\ref{fig:attacker}):
Tendermint would expect the node (belonging to the \emph{victim}) to do
the validation work while the owner
of the private key of the \<Attacker> could just wait for accrued rewards to spend.
A sort of validator's slavery.
Sec.~\ref{sec:shared_entities} asserted that the \emph{Consistency of Shareholders}
property holds, at source level.
Namely, an \<attacker> (of type \<Attacker>) can only become shareholder
by accepting an ongoing sale \<offer> of shares through a call to
\<tv.accept(offer.cost, attacker, offer)> (see Fig.~\ref{fig:simple_shared_entity}).
This is impossible at source level, where that call does \emph{not} compile, since \<attacker> has type \<Attacker> that
is not an instance of \<V>, which has been set to \<TendermintED25519Validator>.
But a Hotmoka blockchain contains only the bytecode of \<SimpleSharedEntity>,
where the signature of \<accept> has been erased into
\<accept(BigInteger amount, PayableContract buyer, Offer offer)>.
Hence a blockchain transaction that invokes \<tv.accept(offer.cost, attacker, offer)>
at bytecode level
\emph{does} succeed, since \<attacker> is an externally owned account and all such accounts
are instances of \<PayableContract> (Fig.~\ref{fig:hierarchy-entities}).
That transaction adds \<attacker> to the shareholders of \<tv>,
therefore violating the \emph{Consistency of Shareholders} property and allowing
validator's slavery.

@@

>>>>

@@

\begin{figure}[b]
\centering
\includegraphics[width=.9\linewidth]{figures/attack}
\caption{Example of possible attack to a smart contract using Java generics.}
\label{figure.attack}
\end{figure}

In particular, it is important that the \emph{Consistency of Shareholders} property holds
for the subclass \<TendermintValidators>: its shareholders
must be \<TendermintED25519Validator>s (as declared in the generic
signature of \<TendermintValidators> in Fig.~\ref{fig:validators}) that enforce a
match between their public key, that identifies who can spend the rewards sent to the validator,
and their Tendermint identifier, that identifies which node of the blockchain
must do the validation work (see how the constructor initializes \<this.id> in
Fig.~\ref{fig:validator}).
If it were possible to add a shareholder of another
type \<Attacker>, the code of \<Attacker> could decouple the node identifier from its
public key (see Fig.~\ref{fig:attacker}):
Tendermint would expect the node (belonging to the \emph{victim}) to do
the validation work while the owner
of the private key of the \<Attacker> could just wait for accrued rewards to spend.
A sort of validator's slavery.
Sec.~\ref{sec:shared_entities} asserted that the \emph{Consistency of Shareholders}
property holds, at source level.
Namely, an \<attacker> (of type \<Attacker>) can only become shareholder
by accepting an ongoing sale \<offer> of shares through a call to
\<tv.accept(offer.cost, attacker, offer)> (Fig.~\ref{fig:simple_shared_entity}).
This is impossible at source level (left part of Fig.~\ref{figure.attack}),
where that call does \emph{not} compile, since \<attacker> has type \<Attacker> that
is not an instance of \<V>, which has been set to \<TendermintED25519Validator>.
But a Hotmoka blockchain contains only the bytecode of \<SimpleSharedEntity>,
where the signature of \<accept> has been erased into
\<accept(BigInteger amount, PayableContract buyer, Offer offer)>
(see Fig.~\ref{figure.java_generics_erasure} and the right part of
Fig.~\ref{figure.attack}).
Hence a blockchain transaction that invokes \<tv.accept(offer.cost, attacker, offer)>
at bytecode level
\emph{does} succeed, since \<attacker> is an externally owned account and all such accounts
are instances of \<PayableContract> (Fig.~\ref{fig:hierarchy-entities}).
That transaction adds \<attacker> to the shareholders of \<tv>,
therefore violating the \emph{Consistency of Shareholders} property and allowing
validator's slavery.

### 2022-03-01 cluster_computing/040_solution.tex
solution figures
Reorganized figures

\section{Fixing the Compilation of the Contract} \label{sec:fix}

@@

\begin{figure}[t]
  \begin{center}
    \begin{lstlisting}[language=Takamaka]
public class TendermintValidators
   extends AbstractValidators<TendermintED25519Validator> {

@@

  public void accept
   (BigInteger,TendermintED25519Validator,Offer)
   aload_0
   aload_1
   aload_2
   aload_3
   invokespecial AbstractValidators
     .accept(BigInteger,PayableContract,Offer)
   return

  // synthetic bridge method
  public void accept(BigInteger,PayableContract,Offer)
   aload_0
   aload_1
   aload_2
   checkcast TendermintED25519Validator
   aload_3
   invokevirtual accept
      (BigInteger,TendermintED25519Validator,Offer)
   return
}
\end{lstlisting}

\noindent
The first \<accept> method above is the compilation of that from Fig.~\ref{fig:solution}:
it delegates to the \<accept> method of the superclass \<AbstractValidators>. The second
\<accept> method above is a \emph{bridge method} that the compiler generates in order to guarantee
that all calls to the erased signature \<accept(BigInteger,PayableContract,Offer)> actually
get forwarded to the first, redefined \<accept>. It casts its \<buyer> argument
into \<TendermintED25519Validator> and calls the first \<accept>. This
bridge method and its checked cast guarantee that only \<TendermintED25519Validator>s
can become validators. For instance, an instance of \<Attacker> (Fig.~\ref{fig:attacker})
cannot be passed to the first \<accept> (type mismatch) and makes the second \<accept>
fail with a class cast exception. The \emph{Consistency of Shareholders} holds
for instances of \<TendermintValidators> now and the attack in Sec.~\ref{sec:attack} cannot occur
anymore.

The solution of redefining method \<accept> can be seen as a limited form of heterogeneous
compilation of generics, restricted to a specific method and forced manually.
It is interesting to consider which methods
would need that redefinition, in general. They are those that have a parameter of a generic type
that is restricted in a subclass. For instance, method \<accept> in
Fig.~\ref{fig:shared_entity} has parameters \<buyer> and \<offer> of generic type
\<S> and \<O>, respectively. The subclass in Fig.~\ref{fig:validators} restricts
\<S> to be a \<TendermintED25519Validator> and \<O> to be
an {\codesize\texttt{Offer<TendermintED25519Validator>}}. Hence we must redefine
\<accept> in the subclass with the more specific types for the \<buyer> and \<offer>
parameters. In the future, a compiler might perform this automatically or a static analysis
tool might issue a warning when such redefinition is needed. Currently, however, this is
left to the programmer of the smart contracts, who might overlook the issue
and give rise to security issues, as shown in Sec.~\ref{sec:attack}.

>>>>

\section{Fixing the Compilation of the Contract}\label{sec:fix}

@@

\begin{figure}[ht]
  \begin{center}
    \begin{lstlisting}[language=Takamaka]
public class TendermintValidators
   extends AbstractValidators<TendermintED25519Validator> {

@@

  public void accept(BigInteger,TendermintED25519Validator,Offer)
   aload_0
   aload_1
   aload_2
   aload_3
   invokespecial AbstractValidators.accept
                        (BigInteger,PayableContract,Offer)
   return

  // synthetic bridge method
  public void accept(BigInteger,PayableContract,Offer)
   aload_0
   aload_1
   aload_2
   checkcast TendermintED25519Validator
   aload_3
   invokevirtual accept(BigInteger,TendermintED25519Validator,Offer)
   return
}
\end{lstlisting}

\noindent
The first \<accept> method above is the compilation of that from Fig.~\ref{fig:solution}:
it delegates to the \<accept> method of the superclass \<AbstractValidators>. The second
\<accept> method above is a \emph{bridge method} that the compiler generates in order to guarantee
that all calls to the erased signature \<accept(BigInteger,PayableContract,Offer)> actually
get forwarded to the first, redefined \<accept>. It casts its \<buyer> argument
into \<TendermintED25519Validator> and calls the first \<accept>. This
bridge method and its checked cast guarantee that only \<TendermintED25519Validator>s
can become validators. As shown in Fig.~\ref{figure.solution},
an instance of \<Attacker> (Fig.~\ref{fig:attacker})
cannot be passed to the first \<accept> (type mismatch) and makes the second \<accept>
fail with a class cast exception. The \emph{Consistency of Shareholders} holds
for instances of \<TendermintValidators> now and the attack in Sec.~\ref{sec:attack} cannot occur
anymore.

The solution of redefining method \<accept> can be seen as a limited form of heterogeneous
compilation of generics, restricted to a specific method and forced manually.
It is interesting to consider which methods
would need that redefinition, in general. They are those that have a parameter of a generic type
that is restricted in a subclass. For instance, method \<accept> in
Fig.~\ref{fig:shared_entity} has parameters \<buyer> and \<offer> of generic type
\<S> and \<O>, respectively. The subclass in Fig.~\ref{fig:validators} restricts
\<S> to be a \<TendermintED25519Validator> and \<O> to be
an {\codesize\texttt{Offer<TendermintED25519Validator>}}. Hence we must redefine
\<accept> in the subclass with the more specific types for the \<buyer> and \<offer>
parameters. In the future, a compiler might perform this automatically or a static analysis
tool might issue a warning when such redefinition is needed. Currently, however, that is
left to the programmer of the smart contracts, who might overlook the problem
and give rise to security issues, as shown in Sec.~\ref{sec:attack}.

\begin{figure}[ht]
\centering
\includegraphics[width=.95\linewidth]{figures/solution}
\caption{Example of how the proposed solution works at run time when the \<accept> method is called by a \<Validator> or an \<Attacker>.}
\label{figure.solution}
\end{figure}
